Allowing Stable Groups of Prosumers to Enter the Market through Proper Aggregation

### 2014-09-12 Article.tex

Traditionally, forming coalitions in a pool of agents can be done either in a centralized way where a single central unit is responsible for all the computations or in a distributed way where agents have only local knowledges and take actions accordingly. It is of common use to represent the situation and assess the stability of the solution by using game theoretic tools. Some papers focus on finding an optimal coalition structure giving a pool of autonomous self interested agents. In this direction, \cite{Saad2009} and \cite{Luan2014} design a distributed merge and split algorithm based on modified Pareto order with utility functions aiming at optimizing a parameter of interest such as energy losses or communication costs.

@@

The setting is thus similar with some fiancial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} Mantegna introduced another approach where stocks, represented by the timeseries of their daily log returns, are organized in a graph such that stocks (vertices) exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph. Because there is a distance between every two nodes, \cite{Mantegna1999} obtains a complete graph where precious information is flooded. In order to discard unsignificant information, \cite{Mantegna1999} used a minimum spanning tree oriented technique and achieves a hierarchical clustering of the stocks. Since then, several other filtering techniques have been proposed, such as k-nearest graphs, or $ \epsilon $-graphs, that relax the tree assumption \cite{Onnela2004}.

The idea of $\epsilon$-graphs consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. Despite the procedure simplicity, the choice of $ \epsilon $ (or k for k-nearest graphs) is not trivial and may influence strongly the results. There is indeed to our knowledge no optimal rule for chosing $ \epsilon $. In \cite{Garas2008} and \cite{Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques...) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. One of the conclusions was that "strong links" (i.e links between strongly correlated timeseries) are responsible for clustering while "weak links" provide network connectivity.

Presented in this way, the timeseries clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweeness \cite{Girvan2002}, spectrum of Laplacian \cite{Newman}...), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and the idea behind it is actually quite simple. It starts indeed by searching for cliques (a complete subgraph) in the given graph as potential community cores. Because other nodes can naturally belong to a community with a less strong connectivity requirement, the algorithm uses a fitness function that quantifies the quality (according to some criterions) of a community structure inside the graph. Expansed cliques alternatively look in their respective neighborhood, select the node that yields the best increase in fitness, and finally incorporate it. The percolation algorithm stops when seeds cease to grow or when all nodes are affected to at least one seed. At this point, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where persons often belongs to several communities (family, friends, colleagues...).

@@

\begin{center}
\begin{figure}
\setlength{\tabcolsep}{-0.7em}
\begin{tabular}{cc}
   \includegraphics[scale=0.5]{./figure1/util.eps} &
   \includegraphics[scale=0.5]{./figure6/util2.eps} \\
\end{tabular}
\caption{Utility function in parameters space. Left in $ (\phi, P_{\phi}^{MIN}) $ when $ N_{COAL} $ is fixed. Right in $ (N_{COAL}, P_{\phi}^{MIN}) $ for a fixed $ \phi $}
\label{Fig2}
\end{figure}
\end{center}

As visible on the left plot of figure \ref{Fig2}, the parameters $\phi$ and $P_{\phi}^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higher utilities.

@@

On the right plot of figure \ref{Fig2}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P_{\phi}^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances also decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of $ P_{\phi}^{MIN} $ for clique percolation (blue dots) and random process (red dots).}
  \label{Fig3}
\end{figure}
\end{center}

@@

In more details, the left plot of figure \ref{Fig4} presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for both random (red curve) and clique percolation (blue curve). As for the right plot, it shows the percentage of coalitions able to enter the market for different values of $ P_{\phi}^{MIN} $. For consistency, we average the results of both plots of the random procedure over 100 realizations (5 realizations for clique percolation, but the results are relatively stable), and the errorbars in the plots stands for the standard deviations of the results.

When the grid requirements are constant (left plot), and the number of desired coalitions is low, clique percolation generally performs only a little better than random. Nevetheless, when $ N_{COAL} $ gets bigger, the performance of a random split tumble down rapidely while clique percolation social welfare decreases slowly.

When the grid requirements vary, for very low $ P_{\phi}^{MIN} $ (close to zero), all coalitions for both algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{\phi}^{MIN} $ increases, we see the percentage of the random procedure quickly dropping while it stays constant for our algorithm. For $ P_{\phi}^{MIN} = 4 $, we see that only a little more than half of the coalitions for the random case are able to enter while all of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes too high, the acceptance percentage of our algorithm tends slowly to zero (only partly shown in this plot for readability).

\begin{center}
\begin{figure}
\setlength{\tabcolsep}{0em}
\begin{tabular}{cc}
   \includegraphics[scale=0.45]{./figure7/uti.eps} &
   \includegraphics[scale=0.45]{./figure4/survival.eps} \\
\end{tabular}
   \caption{Percentage of coalitions able to enter the market for different values of $ P_{MIN} $ for clique percolation (blue) and random process (red).}
   \label{Fig4}
\end{figure}
\end{center}

>>>>

Traditionally, forming coalitions in a pool of agents can be done either in a centralized way where a single central unit is responsible for all the computations or in a distributed way where agents have only local knowledges and take actions accordingly. It is of common use to represent the situation and assess the stability of the solution by using game theoretic tools. Some papers \cite{Saad2009} \cite{Luan2014} focus then on finding an optimal coalition structure giving a pool of autonomous self interested agents using distributed merge/split algorithms.

@@

The setting is thus similar with some fiancial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} Mantegna introduced another approach where stocks, represented by the timeseries of their daily log returns, are organized in a graph such that stocks exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph.

The idea of $\epsilon$-graphs, in this context, consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. In \cite{Garas2008} and \cite{Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques...) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. One of the conclusions was that "strong links" (i.e links between strongly correlated timeseries) are responsible for clustering while "weak links" provide network connectivity.

Presented in this way, the timeseries clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweeness \cite{Girvan2002}, spectrum of Laplacian \cite{Newman}...), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and employs a greedy expansion rule based on a fitness function. When the algorithm stops, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where persons often belongs to several communities (family, friends, colleagues...).

@@

\begin{figure}
 \centering
  \subcaptionbox{$ (\phi, P_{\phi}^{MIN}) $ space\label{fig2:a}}{\includegraphics[width=1.6in]{./figure1/util.eps}}\hspace{1em}%
  \subcaptionbox{$ (N_{COAL}, P_{\phi}^{MIN}) $ space\label{fig2:b}}{\includegraphics[width=1.6in]{./figure6/util2.eps}}
\caption{Utility function in parameters space : In $ (\phi, P_{\phi}^{MIN}) $ when $ N_{COAL} $ is fixed (fig \ref{fig2:a}), and in $ (N_{COAL}, P_{\phi}^{MIN}) $ for a fixed $ \phi $ (fig \ref{fig2:b})}
\label{Fig2}
\end{figure}

As visible on figure \ref{fig2:a}, the parameters $\phi$ and $P_{\phi}^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higher utilities.

@@

On figure \ref{fig2:b}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P_{\phi}^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances also decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of $ P_{\phi}^{MIN} $ for clique percolation (blue dots) and random process (red dots). The diameters of the dots are proportional to the corresponding coalition sizes.}
  \label{Fig3}
\end{figure}
\end{center}

@@

In more details, figure \ref{fig4:a} presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for both random (red curve) and clique percolation (blue curve). As for the right plot, it shows the percentage of coalitions able to enter the market for different values of $ P_{\phi}^{MIN} $. For consistency, we average the results of both plots of the random procedure over 100 realizations (5 realizations for clique percolation, but the results are relatively stable), and the errorbars in the plots stands for the standard deviations of the results.

When the grid requirements are constant (figure \ref{fig4:b}), and the number of desired coalitions is low, clique percolation generally performs only a little better than random. Nevetheless, when $ N_{COAL} $ gets bigger, the performance of a random split tumble down rapidely while clique percolation social welfare decreases slowly.

When the grid requirements vary, for very low $ P_{\phi}^{MIN} $, all coalitions for both algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{\phi}^{MIN} $ increases, we see the percentage of the random procedure quickly dropping while it stays constant for our algorithm. For $ P_{\phi}^{MIN} = 4 $, we see that only a little more than half of the coalitions for the random case are able to enter while all of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes too high, the acceptance percentage of our algorithm tends slowly to zero (only partly shown in this plot for readability).

\begin{figure}
 \centering
  \subcaptionbox{Social Welfare\label{fig4:a}}{\includegraphics[width=1.6in]{./figure7/uti.eps}}\hspace{1em}%
  \subcaptionbox{Market entering percentage\label{fig4:b}}{\includegraphics[width=1.6in]{./figure4/survival.eps}}
\caption{Evolution of the social welfare (fig \ref{fig4:a}) and the percentage of coalitions able to enter the market (fig \ref{fig4:b}) for different values of $ P_{MIN} $. Blue curves represent clique percolation and the red ones, the random process.}
\label{Fig4}
\end{figure}
